---
title: "Assignment 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: "Your Name Here"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

**Learning Objectives:**
- Apply spatial operations to answer policy-relevant research questions
- Integrate census demographic data with spatial analysis
- Create publication-quality visualizations and maps
- Work with spatial data from multiple sources
- Communicate findings effectively for policy audiences

---

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and brief explanations:

#### Step 1: Data Collection (5 points)

Load the required spatial data:
- Pennsylvania county boundaries
- Pennsylvania hospitals (from lecture data)
- Pennsylvania census tracts

**Your Task:**
```{r}
# Load required packages
library(sf)
library(tidyverse)
library(tigris)
library(tidycensus)
library(scales)
library(patchwork)
library(here)
library(knitr)
options(tigris_use_cache = TRUE, tigris_progress = FALSE) 
# Load spatial data
pa_counties <- st_read(here("data/Pennsylvania_County_Boundaries.shp"))
hospitals <- st_read(here("data/hospitals.geojson"))
census_tracts <- tracts(state = "PA", cb = TRUE)

# Check that all data loaded correctly
head(pa_counties)
head(hospitals)
head(census_tracts)
```

**Questions to answer:**
- How many hospitals are in your dataset?

  233

- How many census tracts?

  3445

- What coordinate reference system is each dataset in?

  census tract and metro areas are in NAD83, while the rest three are in WGS84
---

#### Step 2: Get Demographic Data 

Use `tidycensus` to download tract-level demographic data for Pennsylvania.

**Required variables:**
- Total population
- Median household income
- Population 65 years and over (you may need to sum multiple age categories)

**Your Task:**
```{r}
# Get demographic data from ACS
df=load_variables(2023,"acs5/profile")
data=c(population="DP02_0018",age="DP05_0024",income="DP03_0062")
df = get_acs(state = "PA",
              variables = data,
              survey = "acs5",
             geometry = TRUE,
             geography = "tract",
             output = "wide",
             year = 2023)



# Join to tract boundaries
#I already downloaded my geography when using tidycensus
sum(is.na(df$incomeE))
median(df$incomeE,na.rm = TRUE)
```

**Questions to answer:**
- What year of ACS data are you using?

  2019-2023 5 year acs

- How many tracts have missing income data?

  66

- What is the median income across all PA census tracts?

  72943.5
  
---

#### Step 3: Define Vulnerable Populations 

Identify census tracts with vulnerable populations based on TWO criteria:
1. Low median household income (choose an appropriate threshold)
2. Significant elderly population (choose an appropriate threshold)

**Your Task:**
```{r}
# Filter for vulnerable tracts based on your criteria

summary(df)

vulnerable=df%>%
  filter(ageE>945|incomeE<32150)

print(vulnerable)

974/3446

```

**Questions to answer:**

- What income threshold did you choose and why?

  I chose $32,150, which is the penn poverty line 

- What elderly population threshold did you choose and why?

  I chose the 3rd quantile 

- How many tracts meet your vulnerability criteria?

  974

- What percentage of PA census tracts are considered vulnerable by your definition?

  28%


---

#### Step 4: Calculate Distance to Hospitals 

For each vulnerable tract, calculate the distance to the nearest hospital.

**Your Task:**
```{r}

# Transform to appropriate projected CRS
vulnerable_proj=vulnerable%>%
  st_transform(3365)

host_proj=hospitals%>%
  st_transform(3365)

# Calculate distance from each tract centroid to nearest hospital
vulnerable_center=st_centroid(vulnerable_proj)

dist <- vulnerable_center %>%
  mutate(
    dis_to_tract = apply(
      st_distance(st_centroid(vulnerable_center), host_proj),
      1,
      min
    ) *0.000189
  )

mean(dist$dis_to_tract)
max(dist$dis_to_tract)
dist$indi=ifelse(dist$dis_to_tract>15,1,0)
summary(dist$indi==1)

```

**Requirements:**
- Use an appropriate projected coordinate system for Pennsylvania
- Calculate distances in miles
- Explain why you chose your projection

**Questions to answer:**
- What is the average distance to the nearest hospital for vulnerable tracts?
  
  4.17

- What is the maximum distance?
  
  25.91

- How many vulnerable tracts are more than 15 miles from the nearest hospital?
  
  17

---

#### Step 5: Identify Underserved Areas 

Define "underserved" as vulnerable tracts that are more than 15 miles from the nearest hospital.

**Your Task:**
```{r}
# Create underserved variable
underserved=dist%>%
  filter(indi==1)
17/973


```

**Questions to answer:**
- How many tracts are underserved?
  
  17

- What percentage of vulnerable tracts are underserved?
  
  1.7%

- Does this surprise you? Why or why not?
  
  Yes, as 15 miles is such a long way and there are still 17 tracts that are so far away, it should be less than 10
  
---

#### Step 6: Aggregate to County Level

Use spatial joins and aggregation to calculate county-level statistics about vulnerable populations and hospital access.

**Your Task:**
```{r}
# Spatial join tracts to counties
df = get_acs(state = "PA",
              variables = data,
              survey = "acs5",
             geometry = TRUE,
             geography = "county",
             output = "wide")

# Aggregate statistics by county

dist_final=dist%>%
  st_drop_geometry()%>%
  mutate(NAME=str_remove(NAME,"; Pennsylvania"))

dist_final[,11:12] = str_split_fixed(dist_final$NAME,";",2)

df_removed=df%>%
  mutate(NAME=str_remove(NAME,", Pennsylvania"))%>%
  mutate(NAME = str_trim(str_to_title(NAME)))
         
dist_final2=dist_final%>%
  mutate(NAME=V12)%>%
  mutate(NAME = str_trim(str_to_title(NAME)))%>%
  group_by(NAME)%>%
  summarise(
    vulnerable=n(),
    underserved=sum(indi==1),
    per=underserved/vulnerable,
    avg_dis=mean(dis_to_tract),
    vulner_pop=sum(populationE)
  )

df_removed=df_removed%>%
  left_join(dist_final2,by="NAME")

```

**Required county-level statistics:**
- Number of vulnerable tracts
- Number of underserved tracts  
- Percentage of vulnerable tracts that are underserved
- Average distance to nearest hospital for vulnerable tracts
- Total vulnerable population

**Questions to answer:**
- Which 5 counties have the highest percentage of underserved vulnerable tracts?

  Pike, Wyoming, Bradford, Clearfield, and Snyder

- Which counties have the most vulnerable people living far from hospitals?

  Allegheny County

- Are there any patterns in where underserved counties are located?

  They are all far from the major cities, which tends to have better medical care services
---

#### Step 7: Create Summary Table 

Create a professional table showing the top 10 priority counties for healthcare investment.

**Your Task:**
```{r}
library(scales)
# Create and format priority counties table
df_table=df_removed%>%
  st_drop_geometry()%>%
  mutate(percentage=round(per*100,digits = 3))%>%
  mutate(pop=comma(populationE))%>%
  mutate(distance=round(avg_dis,digits = 3))%>%
  select(NAME,pop,vulnerable,underserved, percentage, distance, vulner_pop)
 
final_table=df_table[order(df_table$percentage,decreasing=TRUE),]




kable(final_table, col.names=c("County Name","Population","Number of Vulnerable Tracts","Underserved Tracts",
                            "Percentage of Underserved Tract","Average Distance to Nearest Hospital",
                            "Vulnerable Population"),format = "html",row.names = FALSE,
      bootstrap_options=c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center",caption = "County-level summary of population, vulnerable tracts, underserved tracts, and hospital accessibility in Pennsylvania.")

```

**Requirements:**
- Use `knitr::kable()` or similar for formatting
- Include descriptive column names
- Format numbers appropriately (commas for population, percentages, etc.)
- Add an informative caption
- Sort by priority (you decide the metric)

---

## Part 2: Comprehensive Visualization 

Using the skills from Week 3 (Data Visualization), create publication-quality maps and charts.

### Map 1: County-Level Choropleth 

Create a choropleth map showing healthcare access challenges at the county level.

**Your Task:**
```{r}
# Create county-level access map
df_removed$percentage=df_removed$per*100
ggplot(df_removed) +
  geom_sf(aes(fill = percentage), color = "white")+
  geom_sf(data = host_proj, aes(color = "Hospital"),size=1)+
  scale_fill_viridis_c(
    name = "Percentage of vulnerable tracts",
    option = "magma"
  ) +
  scale_color_manual(
    name="",
    values = c("Hospital" = "red")
  )+
  labs(
    title = "Vulnerable Tracts and Hospital Locations in Pennsylvania",
    subtitle="Data Source: ACS 5-Year Census Data"
  ) +
  theme_void()



```

**Requirements:**
- Fill counties by percentage of vulnerable tracts that are underserved
- Include hospital locations as points
- Use an appropriate color scheme
- Include clear title, subtitle, and caption
- Use `theme_void()` or similar clean theme
- Add a legend with formatted labels

---

### Map 2: Detailed Vulnerability Map 

Create a map highlighting underserved vulnerable tracts.

**Your Task:**
```{r}
# Create detailed tract-level map
dist_final_map=dist_final%>%
  group_by(GEOID)%>%
  summarise(
    vulnerable=n(),
    underserved=sum(indi==1),
    per=underserved/vulnerable,
    avg_dis=mean(dis_to_tract),
    vulner_pop=sum(populationE)
  )

dist_final_map2=vulnerable_proj%>%
  left_join(dist_final_map,by="GEOID")

ggplot() +
  geom_sf(data = census_tracts, color="black")+
  geom_sf(data=dist_final_map2,fill="red",aes(color="Vulnerable Tract"))+
  geom_sf(data = host_proj, aes(color = "Hospital"),size=1)+
  geom_sf(data = df_removed,fill= NA,aes(color = "County Boundary"),size=6, alpha= 1)+
  scale_fill_viridis_c(
    name = "Percentage of vulnerable tracts",
    option = "magma"
  ) +
  scale_color_manual(
    name="",
    values = c("Hospital" = "black","Vulnerable Tract"="black","County Boundary"="yellow")
  )+
  labs(
    title = "Vulnerable Tracts and Hospital Locations in Pennsylvania",
    subtitle = "Data Source: ACS 5-Year Census Data"
  ) +
  theme_void()
```

**Requirements:**
- Show underserved vulnerable tracts in a contrasting color
- Include county boundaries for context
- Show hospital locations
- Use appropriate visual hierarchy (what should stand out?)
- Include informative title and subtitle

---

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to hospitals for vulnerable populations.

**Your Task:**
```{r}
# Create distribution visualization
ggplot(dist_final, aes(x = dis_to_tract,weight=populationE)) +
  geom_histogram(binwidth = 1, color = "black", fill = "skyblue") +
   scale_y_continuous(labels = comma) +
  labs(
    title = "Distribution of Distances to Hospitals for Population",
    x = "Distance (Miles)",
    y = "Population",
    subtitle = "The distribution is left-skewed, with most people located relatively close to the nearest hospital."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12)
  )


```

**Suggested chart types:**
- Histogram or density plot of distances
- Box plot comparing distances across regions
- Bar chart of underserved tracts by county
- Scatter plot of distance vs. vulnerable population size

**Requirements:**
- Clear axes labels with units
- Appropriate title
- Professional formatting
- Brief interpretation (1-2 sentences as a caption or in text)

---

## Part 3: Bring Your Own Data Analysis 

### Your Analysis

**Your Task:**

1. **Find and load additional data**
   - Document your data source
   - Check and standardize the CRS
   - Provide basic summary statistics

```{r}
# Load your additional dataset
library(httr)
library(jsonlite)

tree=st_read("C:/Users/wensh/Desktop/MUSA5080/week-04/data/ppr_tree_inventory_2024/ppr_tree_inventory_2024.shp")
parkl=st_read("C:/Users/wensh/Desktop/MUSA5080/week-04/data/PPR_Program_Sites.geojson")

park_proj=parkl%>%st_transform(3365)
tree_proj=tree%>%st_transform(3365)
summary(park_proj)
summary(tree_proj)
```

**Questions to answer:**
- What dataset did you choose and why?

  I choose both tree and park data in Philadelphia because I want to study if different tracts have equal access to green spaces

- What is the data source and date?

  Both of them are from Opendataphilly, tree data is from 2024 while park data is from 2022

- How many features does it contain?

  there are 151,713 features in tree data and 171 features in park data

- What CRS is it in? Did you need to transform it?

  Both of the CRS are WGS84 at first and I transform it to espg 3365

---

2. **Pose a research question**

Write a clear research statement that your analysis will answer.

**Do minority and low income census tracts have the same access to green spaces and enjoy same tree canopy as others?**

---


3. **Conduct spatial analysis**

Use at least TWO spatial operations to answer your research question.

**Required operations (choose 2+):**
- Buffers
- Spatial joins
- Spatial filtering with predicates
- Distance calculations
- Intersections or unions
- Point-in-polygon aggregation

**Your Task:**
```{r}
# Your spatial analysis
#get philly census data
philly=get_acs(
  geography = "tract",
  state = "PA",
  county = "101",
  year = 2023,
  survey = "acs5",
  variables =c(population="DP02_0018",age="DP05_0023",income="DP03_0062"),
  geometry = TRUE,
  output="wide"
)
#reproject census data
philly_project=philly%>%
  st_transform(3365)
#spatial join census data with tree counts
philly_tree = philly_project%>%
  mutate(tree_count = lengths(st_contains(., tree_proj)))

#calculate the distance between tract and parks
tract_center=st_centroid(philly_project)
dist_park <- tract_center %>%
  mutate(
    dis_to_tract = apply(
      st_distance(tract_center, park_proj),
      1,
      min
    ) *0.000189
  )

#merge tree counts and distance to park
dist_final_philly = dist_park%>%
  st_drop_geometry()%>%
  select(GEOID,dis_to_tract)
final_philly = philly_tree%>%
  left_join(dist_final_philly,by="GEOID")

#calculate tree per area in census tract
final_philly = final_philly%>%
  mutate(area=st_area(geometry))%>%
  mutate(tree_per=as.numeric(tree_count/area))
  
#normalize both distance to park and tree per area
normalize_to_0_1 <- function(vec) {
  min_val <- min(vec, na.rm = TRUE)
  max_val <- max(vec, na.rm = TRUE)
  normalized_vec <- (vec - min_val) / (max_val - min_val)
  return(normalized_vec)
}

#calculate the final index (park 75% weight and tree 25% weight)
final_philly=final_philly%>%
  mutate(normalize_tree=normalize_to_0_1(tree_per))%>%
  mutate(normalize_park=normalize_to_0_1(dis_to_tract))%>%
  mutate(index=0.75*normalize_park+0.25*normalize_tree)

#map the results
ggplot(final_philly) +
  geom_sf(aes(fill = index), color = "white")+
  scale_fill_viridis_c(
    name = "Green Space Index",
    option = "magma"
  ) +
  labs(
    title = "Green Space Index in Pennsylvania",
    subtitle = "Calculated based on distance to park and tree by area"
  ) +
  theme_void()
```

**Analysis requirements:**
- Clear code comments explaining each step
- Appropriate CRS transformations
- Summary statistics or counts
- At least one map showing your findings
- Brief interpretation of results (3-5 sentences)

**Your interpretation:**

Based on the map, it is clear that Richer areas (eg. Center City) obtain a higher green space index, while poor neighborhoods (especially in the north such as Kensington) have a low green space index. As a result, minority and poor neighborhood indeed have less access to green space. 

Another finding is suburban area tend to have larger green space index compared to downtown, which makes sense as there are more spaces to plant trees/parks at suburb comapred to downtown.


## Finally - A few comments about your incorporation of feedback!

First I remove the extra instruction text in my homework such as the submission requirement in the end, I also remove the [] in my answer. 

Second, I used round() function to make my table look cleaner



